The 2018 heatwave and its implications on ozone induced damage on vegetation in a subarctic climate

### 2020-04-07 VegOzone.tex
Ysterdays text spell-check.

\introduction  %% \introduction[modified heading if necessary]
\label{sec:intro}
\begin{itemize}
\item ozone in general
\item ozone induced damage in vegetation \emph{Contributor: Ane - maybe also ask Aud?}
  Ozone acts as an oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves.Ozone enters the leaves through stomata, that are kept open to allow for gas exchange between the leaf air spaces and the exterior air. When the stomata are closed, ozone does not enter.
\item metrices of ozone: aot40 and pod
\item present subarctic climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] midnight sun and 23\,\unit{h} open stomata (biological details will be published in a component paper)
  \item[$\rightarrow$] ozone climatology and spring peak
  \end{itemize}
\item future climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] increase in heatwaves and dry periods
  \item[$\rightarrow$] shift in start of growing season (into vicinity of ozone spring peak)
  \item[$\rightarrow$] increase in tropospheric background ozone
  \end{itemize}
\item meteorological situation 2018 vs 2019
\end{itemize}

@@

\subsection{NIBIO research station Svanhovd}
\label{subsec:svanhovd}
"Svanhovd is located beside the Pasvik river, in the middle of the wedge of Norwegian land separating Russia and Finland in the north. It is where the eastern Siberian taiga meets the western Boreal forest. [...] Svanhovd was originally established as a State Demonstration and Experimental Farm in the 1930's, in order to expand agriculture in Pasvik valley [...]" - \href{https://www.nibio.no/en/about-eng/addresses/northern-norway/svanhovd}{NIBIO webpage, accessed March 2020}.\\
The research station at Svanhovd comprises long-term observations of various agrameteorological variables, i.a. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths.

@@

     NIBIO Svanhovd in Pasvik valley is strongly committed to environmental research cooperation in the Barents region. The station disseminates knowledge and performs R\&D based upon the natural environment of this region. It also runs an up-to date DNA-laboratory, a national park centre, a botanical garden and a conference centre. Svanhovd has extensive cooperation with all countries in the Barents region."  - \href{https://www.nibio.no/en/about-eng/addresses/northern-norway/svanhovd}{NIBIO webpage, accessed April 2020}.\\

@@

All agrameterological variables including temperature and precipitation are available from September 1992 to present day (LandbruksMeteorologisk Tjeneste provided by NIBIO -- \href{https://lmt.nibio.no/}{lmt.nibio.no}, note the station name here is Pasvik).

@@

\subsubsection{Ozone garden}
\label{subsubsec:ozone_garden}
      {\bf TODO: Details about the ozone garden and observed damage on ozone sensitive plants. We will need to come back to this in the discussion section after the Do3se modelling part.} \emph{Author: Ane/Aud -- pictures are welcome, just upload them on overleaf or send them to me.}

\subsection{Other data}
\label{subsec:other_data}
To determine the significance of the ozone conditions in 2018, we have to derive a surface ozone concentration climatology for Svanvik. From 1986--1996 ozone monitoring has been operated by NILU at Svanhovd, but no ozone measurements have been conducted for the past $14\,\unit{years}$. Due to probabel changes in northern hemispheric tropospheric background ozone and a reduction of summertime peak values related to ozone precursors that fell under airquality regulations {\bf TODO: Citations!}, we also have to have a look at ozone observations at other sites in northern Fennoscandia. In regard to ozone, available observational data are sparse in this region. An overview over the geographical locations of conducted atmospheric monitoring in the past and present is given in Fig.~\ref{fig:station_map_fennoscandia}. The data is descibed in Section~\ref{subsubsec:ebas}. 
As additional complement to observational data which is not available at all times and spatially sparse, we also look at verious ozone reanalysis products (Section~\ref{subsubsec:ozone_rea}).

@@

\subsubsection{Surface ozone observations}
\label{subsubsec:ebas}
All surface ozone observations used in this study are available from the EBAS database operated by NILU (\href{http://ebas.nilu.no/}{ebas.nilu.no}). In Table~\ref{tab:ebas_obs}, we give an overview over the exact locations and time range of operational data taking. All stations with available long term observation are locate on higher ground than Svanhovd which will result in higher surface ozone concentrations compared to data taken at Svanvik as ozone abundance decreases with decreasing altitude in the absence of local precursors. {\bf TODO: Citation!} {\bf TODO: The following has to be confirmed and ellaborated on!} The vegetation at all sites is similar and consists mainly of pine forests, birch, and heath.

Data from before the 1990s is available solely from Svanvik and Karasjok. These data, however, do not follow the high quality standard procedures implemented nowadays and have to be treated with care {\bf TODO: citation NILU ozone report 200x}. Especially, ozone monitors did not undergo a regular re-calibration. This leads likely to drifts in the observed data and may impose a false periodicity. In 1997, the ozone monitoring at Jergul was relocated downstream the river Karasjokka to a site close by Karasjok. Although the same equipment has been used, this might have intruduced a shift in the data series denotes as Jergul/Karasjok in the following. Only few data have been taken upstream the Pasvik river at Janiskovski, hence these will not be taken into further consideration. As can be seen in Fig.~\ref{fig:ozone_timesseries_fenoscandic_obs}, there is an overlapp of about $4\,\unit{years}$ between data taken at Esrange, Jergul/Karasjok, and Svanvik. The hatched areas herin indicate the time periods with insufficient quality control as mentioned above.

@@

Details of renalysis data sets refered to in this work are listed in Table~\ref{tab:ozone_rea}.
There are two global reanalysis products available at the European Centre for Medium-Range Weather Forecasts (ECMWF) that include atmospheric tracers, in particular ozone (MACC and CAMSRA). The temporal as well as spatial resolution of these reanalysis products is rather coarse: 3-hourly and $0.75^\circ\times 0.75^\circ$ or roughly $78.1\,\unit{km}\times 83.4\,\unit{km}$ at the location of Svanvik, respectively. In addition, higher resolution surface ozone reanalysis for whole Europe is available from Copernicus service for regional air quality control. The regional model reanalysis ensemble mean is at fairly high spatial and temporal resolution compared to the data from ECMWF: $0.1^\circ\times 0.1^\circ$ or roughly $10.4\,\unit{km}\times 11.1\,\unit{km}$ and 1-hourly, respectively. The covered time periods differ and no data are available for before the turn of the millenium. The CAMSRA and CAMS Regional Air Quality ensemble mean are available in near realtime, but only former covers a period of sufficient length for climate analysis (2003 -- present). Their quality with respect to observations will be discussed in Section~\ref{subsubsec:clim_ozone}. Without too much foreshadowing, the results prevented us from fetching the whole CAMSRA data.

@@

\subsubsection{Temperature, precipitation, and global irradiance}
\label{subsubsec:clim_temp_prec}
The location of Svanvik at $69.45\,\unit{^\circ N}$ suggests a subarctic climate. The derived climate diagram following Walter and Lieth shown in Fig.~\ref{fig:climatediagram} supports this very well. The diagram is based on climatological data from Svanvik (1992--2012) downloaded from LandbruksMeteorologiske Tjeneste. Monthly averaged temperatures (red line) are displayed with standard error of mean as error band. Averaged monthly accumulated precipitation (blue bars) is shown with standard deviation of mean as error bars. Months with average temperatures below freezing are denoted with a star. As can be seen, temperatures stay below freezing for 5 consequative months, while only 2 months breach $10\,\unit{^\circ C}$ regularly (July, August), satisfying the conditions for K\"{o}ppens climate classifcation of a regular subarctic climate (Dfc). The highest monthly average temperature is $(13.1\pm 1.1)\,\unit{^\circ C}$ in July and the lowest $(-12.8\pm 2.0)\,\unit{^\circ C}$ in February. In the time period 1992--2012, the coldest measured temperature was $-45.2\,\unit{^\circ C}$ observed January 27, 1999, while the highest temperature on July 16 the same year was $29.4\,\unit{^\circ C}$.

Winter and spring (November--April, except for March) are relatively dry ($\sum_m \mathrm{Precip} < 20\,\unit{mm}$). The driest month is January with $\sum_m \mathrm{Precip} = (16.7\pm 3.0)\,\unit{mm}$. The most precipitation occurres in the summer months, with a highest monthly averaged accumulated precipitation $\sum_m \mathrm{Precip} = (58.5\pm 9.2)\,\unit{mm}$ in August. The average annual accumulated precipitation given with standard error of mean is $(383\pm 86)\,\unit{mm}$. Precipitation shows a large veriablity throughout the years.

@@

\subsubsection{Ozone}
\label{subsubsec:clim_ozone}
In the following, we will look at derived ozone climatologies from different perspectives.
The multi-annual means of daily mean ozone, for simplicity here without standard deviations, at observation sites in northern Fennoscandia (refer to Table~\ref{tab:ebas_obs} for details) are shown in Fig.~\ref{fig:ozone_climatology_fenoscandic_obs}. All climatologies peak in spring, with peak values of about $48\,\unit{ppb}$ in late April. The annual average ozone concentration \chem{\left<[O_3]\right>} declines throughout the growing season until it reaches its minimum in late summer/fall ($24-30\,\unit{ppb}$). The turning points lie approximately in June and November. The ozone climatology laggs about 3 months behind the temperature climatology. At Svanvik, \chem{\left<[O_3]\right>} is $6.6\,\unit{ppb}$ lower than on sites at Jergul/Karasjok (NOR), Esrange (SWE), and Pallas (FIN). The large spread in case of Svanvik and Janiskoski can be explained due to the lower overall statistics.

@@

The correlation of \chem{[O_3]} between Esrange, Pallas, and Jergul/Karasjok is fairly high ($r^2\approx 0.8$, Fig.~\ref{fig:density_distribution}a, b), hence we can put those data series together to derive a general ozone climatology for northern Fennoscandia. In Figure~\ref{fig:ozone_climatology_fenoscandic_obs_spline}, the 2 dimensional density distributions for the thus defined northern Fennoscandia (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}a) and Svanvik (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) are shown. Darker colors indicate higher prohablity to observe these values. On top of the density distributions, $10\,\unit{days}$ averages of daily mean \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, daily maximum \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}, and daily minimum \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} are displayed together with the $1 \sigma$ uncertainties, respectively. In addition, \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} is also shown with standard error of mean (albeit small and hardly visible). Splines have been fitted through the data for easier intercomparison with ranalyis data.

From the northern Fennoscandic climatology, one can deduce several features. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} peaks about $20\,\unit{days}$ earlier than the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}. With respect of the periodicity of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, it takes about $8.5\,\unit{months}$ from minimum concentration to maximum concentration, while from maximum to minimum it only takes about $4.5\,\unit{months}$. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} is at $52\,\unit{ppb}$ in late April and the minimum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} ocurres at $15\,\unit{ppb}$ in the beginning of September. The steep rise of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} in spring coincident with increasing temperatures as well as increase in global irradiance. The decrease in \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} coincides with the estimated onset of the greening season based on the agricultural rule of thumb that the daily average temperature has to be above $5\,\unit{^\circ C}$ for $5\,\unit{days}$ conseguatively (day of the year 110--150, Fig~\ref{fig:greening_season_change_Svanvik}).

As mentioned above, the average ozone concentration at Svanvik is in general lower than at the other sites which is in league with a decrease in tropospheric ozone with decreasing altitude. The densitiy distribution for Svanvik and the respective \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) display a similar pattern as the northern Fennoscandic climatology. Only does the spring peak apparently occure slightly earlier (in the beginning of April).

@@

Finally, we shall look at the comparison of climatologies derived from observational data ozone and the various reanalysis products (Fig.~\ref{fig:ozone_climatologies}). As representative for the sites, the nearest neighboring grid point has been chosen. The ECMWF products do not reproduce observed surface ozone well. The MACC reanalysis is in general too low and displays a erroneous seasonal cycle. The CAMS reanalysis matches better but does not reproduce the seasonality in northern Fennoscandia. The ensemble mean of the Copernicus regional model reanalysis does reproduce the seasonal cycle in northern Fennoscandia, though the modelled ozone concentrations in the north are slightly to low.

\begin{figure*}[t]
  \includegraphics[width=12cm]{ozone_climatologies}
  \caption{Comparison of climatologies derived from observational data (EBAS, see Fig.~\ref{fig:ozone_climatology_fenoscandic_obs}) with various ozone reanalysis products: (a) ECMWF: MACC reanalysis; (b) ECMWF: CAMS reanalysis; (c) Copernicus: Regional model reanalysis ensemble mean. As representative for the sites, the nearest neighboring grid point has been chosen. The ECMWF products do not reproduce observed surface ozone well. The MACC reanalysis is in general too low and displays a erroneous seasonal cycle. The CAMS reanalysis matches better but does not reproduce the seasonality in northern Fennoscandia. The ensemble mean of the Copernicus regional model reanalysis does reproduce the seasonal cycle in northern Fennoscandia, though the modelled ozone concentrations in the north are slightly to low.}
  \label{fig:ozone_climatologies}
\end{figure*}

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{global_rad_signific}
  \caption{Global irradiance at Svanvik. Percentage of days above/below $\pm 1\,\sigma$ from climatological mean displayed for each month. Negative deviations from climatology are shown as negative percentage. The annual positive/negative deviations are indicated in the respective corners (right upper/right lower). The dashed lines indicate the expected percentage of values falling above/below $\pm 1\,\sigma$ if a normal distribution is assumed ($15.9\,\unit{\%}$). Year displayed in panel (a) 2018; (b) 2019. In summer 2018 (May/July), gobal irradiance was significantly higher than on average, while 2019 was generally an average year.}
  \label{fig:global_rad_signific}
\end{figure}

>>>>

\introduction  %% \introduction[modified heading if necessary]
\label{sec:intro}
\begin{itemize}
\item ozone in general
\item ozone induced damage in vegetation \emph{Contributor: Ane - maybe also ask Aud?}
  Ozone acts as an oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves.Ozone enters the leaves through stomata, that are kept open to allow for gas exchange between the leaf air spaces and the exterior air. When the stomata are closed, ozone does not enter.
\item metrics of ozone: aot40 and pod
\item present subarctic climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] midnight sun and 23\,\unit{h} open stomata (biological details will be published in a component paper)
  \item[$\rightarrow$] ozone climatology and spring peak
  \end{itemize}
\item future climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] increase in heatwaves and dry periods
  \item[$\rightarrow$] shift in start of growing season (into vicinity of ozone spring peak)
  \item[$\rightarrow$] increase in tropospheric background ozone
  \end{itemize}
\item meteorological situation 2018 vs 2019
\end{itemize}

@@

\subsection{NIBIO research station Svanhovd}
\label{subsec:svanhovd}
"Svanhovd is located beside the Pasvik river, in the middle of the wedge of Norwegian land separating Russia and Finland in the north. It is where the eastern Siberian taiga meets the western Boreal forest. [...] Svanhovd was originally established as a State Demonstration and Experimental Farm in the 1930's, in order to expand agriculture in Pasvik valley [...]" - \href{https://www.nibio.no/en/about-eng/addresses/northern-norway/svanhovd}{NIBIO web page, accessed March 2020}.\\
The research station at Svanhovd comprises long-term observations of various agrameteorological variables, i.a. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths.

@@

     NIBIO Svanhovd in Pasvik valley is strongly committed to environmental research cooperation in the Barents region. The station disseminates knowledge and performs R\&D based upon the natural environment of this region. It also runs an up-to date DNA-laboratory, a national park centre, a botanical garden and a conference centre. Svanhovd has extensive cooperation with all countries in the Barents region."  - \href{https://www.nibio.no/en/about-eng/addresses/northern-norway/svanhovd}{NIBIO web page, accessed April 2020}.\\

@@

All agrameteorological variables including temperature and precipitation are available from September 1992 to present day (LandbruksMeteorologisk Tjeneste provided by NIBIO -- \href{https://lmt.nibio.no/}{lmt.nibio.no}, note the station name here is Pasvik).

@@

\subsubsection{Ozone garden}
\label{subsubsec:ozone_garden}
      {\bf TODO: Details about the ozone garden and observed damage on ozone sensitive plants. We will need to come back to this in the discussion section after the Do3se modeling part.} \emph{Author: Ane/Aud -- pictures are welcome, just upload them on overleaf or send them to me.}

\subsection{Other data}
\label{subsec:other_data}
To determine the significance of the ozone conditions in 2018, we have to derive a surface ozone concentration climatology for Svanvik. From 1986--1996 ozone monitoring has been operated by NILU at Svanhovd, but no ozone measurements have been conducted for the past $14\,\unit{years}$. Due to probable changes in northern hemispheric tropospheric background ozone and a reduction of summertime peak values related to ozone precursors that fell under air quality regulations {\bf TODO: Citations!}, we also have to have a look at ozone observations at other sites in northern Fennoscandia. In regard to ozone, available observational data are sparse in this region. An overview over the geographical locations of conducted atmospheric monitoring in the past and present is given in Fig.~\ref{fig:station_map_fennoscandia}. The data is described in Section~\ref{subsubsec:ebas}. 
As additional complement to observational data which is not available at all times and spatially sparse, we also look at various ozone reanalysis products (Section~\ref{subsubsec:ozone_rea}).

@@

\subsubsection{Surface ozone observations}
\label{subsubsec:ebas}
All surface ozone observations used in this study are available from the EBAS database operated by NILU (\href{http://ebas.nilu.no/}{ebas.nilu.no}). In Table~\ref{tab:ebas_obs}, we give an overview over the exact locations and time range of operational data taking. All stations with available long term observation are locate on higher ground than Svanhovd which will result in higher surface ozone concentrations compared to data taken at Svanvik as ozone abundance decreases with decreasing altitude in the absence of local precursors. {\bf TODO: Citation!} {\bf TODO: The following has to be confirmed and elaborated on!} The vegetation at all sites is similar and consists mainly of pine forests, birch, and heath.

Data from before the 1990s is available solely from Svanvik and Karasjok. These data, however, do not follow the high quality standard procedures implemented nowadays and have to be treated with care {\bf TODO: citation NILU ozone report 200x}. Especially, ozone monitors did not undergo a regular re-calibration. This leads likely to drifts in the observed data and may impose a false periodicity. In 1997, the ozone monitoring at Jergul was relocated downstream the river Karasjokka to a site close by Karasjok. Although the same equipment has been used, this might have introduced a shift in the data series denotes as Jergul/Karasjok in the following. Only few data have been taken upstream the Pasvik river at Janiskovski, hence these will not be taken into further consideration. As can be seen in Fig.~\ref{fig:ozone_timesseries_fenoscandic_obs}, there is an overlap of about $4\,\unit{years}$ between data taken at Esrange, Jergul/Karasjok, and Svanvik. The hatched areas herein indicate the time periods with insufficient quality control as mentioned above.

@@

Details of reanalysis data sets referred to in this work are listed in Table~\ref{tab:ozone_rea}.
There are two global reanalysis products available at the European Centre for Medium-Range Weather Forecasts (ECMWF) that include atmospheric tracers, in particular ozone (MACC and CAMSRA). The temporal as well as spatial resolution of these reanalysis products is rather coarse: 3-hourly and $0.75^\circ\times 0.75^\circ$ or roughly $78.1\,\unit{km}\times 83.4\,\unit{km}$ at the location of Svanvik, respectively. In addition, higher resolution surface ozone reanalysis for whole Europe is available from Copernicus service for regional air quality control. The regional model reanalysis ensemble mean is at fairly high spatial and temporal resolution compared to the data from ECMWF: $0.1^\circ\times 0.1^\circ$ or roughly $10.4\,\unit{km}\times 11.1\,\unit{km}$ and 1-hourly, respectively. The covered time periods differ and no data are available for before the turn of the millennium. The CAMSRA and CAMS Regional Air Quality ensemble mean are available in near real time, but only former covers a period of sufficient length for climate analysis (2003 -- present). Their quality with respect to observations will be discussed in Section~\ref{subsubsec:clim_ozone}. Without too much foreshadowing, the results prevented us from fetching the whole CAMSRA data.

@@

\subsubsection{Temperature, precipitation, and global irradiance}
\label{subsubsec:clim_temp_prec}
The location of Svanvik at $69.45\,\unit{^\circ N}$ suggests a subarctic climate. The derived climate diagram following Walter and Lieth shown in Fig.~\ref{fig:climatediagram} supports this very well. The diagram is based on climatological data from Svanvik (1992--2012) downloaded from LandbruksMeteorologiske Tjeneste. Monthly averaged temperatures (red line) are displayed with standard error of mean as error band. Averaged monthly accumulated precipitation (blue bars) is shown with standard deviation of mean as error bars. Months with average temperatures below freezing are denoted with a star. As can be seen, temperatures stay below freezing for 5 consecutive months, while only 2 months breach $10\,\unit{^\circ C}$ regularly (July, August), satisfying the conditions for K\"{o}ppens climate classification of a regular subarctic climate (Dfc). The highest monthly average temperature is $(13.1\pm 1.1)\,\unit{^\circ C}$ in July and the lowest $(-12.8\pm 2.0)\,\unit{^\circ C}$ in February. In the time period 1992--2012, the coldest measured temperature was $-45.2\,\unit{^\circ C}$ observed January 27, 1999, while the highest temperature on July 16 the same year was $29.4\,\unit{^\circ C}$.

Winter and spring (November--April, except for March) are relatively dry ($\sum_m \mathrm{Precip} < 20\,\unit{mm}$). The driest month is January with $\sum_m \mathrm{Precip} = (16.7\pm 3.0)\,\unit{mm}$. The most precipitation occurs in the summer months, with a highest monthly averaged accumulated precipitation $\sum_m \mathrm{Precip} = (58.5\pm 9.2)\,\unit{mm}$ in August. The average annual accumulated precipitation given with standard error of mean is $(383\pm 86)\,\unit{mm}$. Precipitation shows a large variability throughout the years.

@@

\subsubsection{Ozone}
\label{subsubsec:clim_ozone}
In the following, we will look at derived ozone climatologies from different perspectives.
The multi-annual means of daily mean ozone, for simplicity here without standard deviations, at observation sites in northern Fennoscandia (refer to Table~\ref{tab:ebas_obs} for details) are shown in Fig.~\ref{fig:ozone_climatology_fenoscandic_obs}. All climatologies peak in spring, with peak values of about $48\,\unit{ppb}$ in late April. The annual average ozone concentration \chem{\left<[O_3]\right>} declines throughout the growing season until it reaches its minimum in late summer/fall ($24-30\,\unit{ppb}$). The turning points lie approximately in June and November. The ozone climatology lags about 3 months behind the temperature climatology. At Svanvik, \chem{\left<[O_3]\right>} is $6.6\,\unit{ppb}$ lower than on sites at Jergul/Karasjok (NOR), Esrange (SWE), and Pallas (FIN). The large spread in case of Svanvik and Janiskoski can be explained due to the lower overall statistics.

@@

The correlation of \chem{[O_3]} between Esrange, Pallas, and Jergul/Karasjok is fairly high ($r^2\approx 0.8$, Fig.~\ref{fig:density_distribution}a, b), hence we can put those data series together to derive a general ozone climatology for northern Fennoscandia. In Figure~\ref{fig:ozone_climatology_fenoscandic_obs_spline}, the 2 dimensional density distributions for the thus defined northern Fennoscandia (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}a) and Svanvik (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) are shown. Darker colors indicate higher probability to observe these values. On top of the density distributions, $10\,\unit{days}$ averages of daily mean \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, daily maximum \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}, and daily minimum \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} are displayed together with the $1 \sigma$ uncertainties, respectively. In addition, \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} is also shown with standard error of mean (albeit small and hardly visible). Splines have been fitted through the data for easier inter-comparison with reanalysis data.

From the northern Fennoscandic climatology, one can deduce several features. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} peaks about $20\,\unit{days}$ earlier than the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}. With respect of the periodicity of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, it takes about $8.5\,\unit{months}$ from minimum concentration to maximum concentration, while from maximum to minimum it only takes about $4.5\,\unit{months}$. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} is at $52\,\unit{ppb}$ in late April and the minimum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} occurs at $15\,\unit{ppb}$ in the beginning of September. The steep rise of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} in spring coincident with increasing temperatures as well as increase in global irradiance. The decrease in \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} coincides with the estimated onset of the greening season based on the agricultural rule of thumb that the daily average temperature has to be above $5\,\unit{^\circ C}$ for $5\,\unit{days}$ consecutively (day of the year 110--150, Fig~\ref{fig:greening_season_change_Svanvik}).

As mentioned above, the average ozone concentration at Svanvik is in general lower than at the other sites which is in league with a decrease in tropospheric ozone with decreasing altitude. The density distribution for Svanvik and the respective \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) display a similar pattern as the northern Fennoscandic climatology. Only does the spring peak apparently occur slightly earlier (in the beginning of April).

@@

Finally, we shall look at the comparison of climatologies derived from observational data ozone and the various reanalysis products (Fig.~\ref{fig:ozone_climatologies}). As representative for the sites, the nearest neighboring grid point has been chosen. The ECMWF products do not reproduce observed surface ozone well. The MACC reanalysis is in general too low and displays a erroneous seasonal cycle. The CAMS reanalysis matches better but does not reproduce the seasonality in northern Fennoscandia. The ensemble mean of the Copernicus regional model reanalysis does reproduce the seasonal cycle in northern Fennoscandia, though the modeled ozone concentrations in the north are slightly to low.

\begin{figure*}[t]
  \includegraphics[width=12cm]{ozone_climatologies}
  \caption{Comparison of climatologies derived from observational data (EBAS, see Fig.~\ref{fig:ozone_climatology_fenoscandic_obs}) with various ozone reanalysis products: (a) ECMWF: MACC reanalysis; (b) ECMWF: CAMS reanalysis; (c) Copernicus: Regional model reanalysis ensemble mean. As representative for the sites, the nearest neighboring grid point has been chosen. The ECMWF products do not reproduce observed surface ozone well. The MACC reanalysis is in general too low and displays a erroneous seasonal cycle. The CAMS reanalysis matches better but does not reproduce the seasonality in northern Fennoscandia. The ensemble mean of the Copernicus regional model reanalysis does reproduce the seasonal cycle in northern Fennoscandia, though the modeled ozone concentrations in the north are slightly to low.}
  \label{fig:ozone_climatologies}
\end{figure*}

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{global_rad_signific}
  \caption{Global irradiance at Svanvik. Percentage of days above/below $\pm 1\,\sigma$ from climatological mean displayed for each month. Negative deviations from climatology are shown as negative percentage. The annual positive/negative deviations are indicated in the respective corners (right upper/right lower). The dashed lines indicate the expected percentage of values falling above/below $\pm 1\,\sigma$ if a normal distribution is assumed ($15.9\,\unit{\%}$). Year displayed in panel (a) 2018; (b) 2019. In summer 2018 (May/July), global irradiance was significantly higher than on average, while 2019 was generally an average year.}
  \label{fig:global_rad_signific}
\end{figure}
